---
title: "Using the `sager` package to import and analyse `sage` results in R"
author:
  - name: Laurent Gatto
    affiliation:
    - Computational Biology and Bioinformatics, UCLouvain
    email: laurent.gatto@uclouvain.be
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('sager')`"
vignette: >
  %\VignetteIndexEntry{The sager package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
    )
suppressPackageStartupMessages(library("BiocStyle"))
```


# Introduction

## The `sager` package

[Sage](https://lazear.github.io/sage/) is a new cross-platform,
extremely performant, open source proteomics search engine [Michael
Lazear](https://github.com/lazear). It uses mzML files as input, is
parameterised by a [json
file](https://github.com/lazear/sage/blob/master/DOCS.md) and is
executed from the command line. In addition to peptide identification,
sage also includes a variety of advanced features such as retention
time prediction and quantification (both isobaric & LFQ). It produces
two files: `results.sage.tsv` with the identification results and,
when configured to quantify features, `quant.tsv`.

The `sager` package uses these two files as input. It facilitates
their import into established Bioconductor classes:

- Identification results are parsed and imported as
  [PSMatch::PSM()](https://rformassspectrometry.github.io/PSMatch/)
  objects with `sagePSM()`.

- Quantitation (and identification) results are parsed, merged and
  imported as
  [QFeatures::QFeatures()](https://rformassspectrometry.github.io/QFeatures/articles/QFeatures.html)
  objects with `sageQFeatures()`.

- Sage results can also be integrated with a
  [Spectra::Spectra()](https://rformassspectrometry.github.io/Spectra/)
  instance (holding the raw data) into a
  [MsExperiment::MsExperiment()](https://rformassspectrometry.github.io/MsExperiment/)
  object, as documented in this document.

The goal of the `sager` package can thus be summarised as a way to
leverage the existing [R for Mass
Spectrometry](https://www.rformassspectrometry.org/) and
[Bioconductor](https://bioconductor.org/) infrastructure to analyse
sage results.

## Installation instructions

To install `sager` and its dependencies, you'll need the `BiocManager`
package, that can be installed from CRAN (unless you already have it):

```{r installbiocmanager, eval = FALSE}
if (!requireNamespace("BiocManager"))
    install.packages("BiocManager")
```

Note that `sager` is under constant development, and thus will depend
on some under-development dependencies, that haven't made it into the
stable releases yet. Here are the instructions to install these:

To install the *devel* version of `PSMatch`:

```{r installpsmatch, eval = FALSE}
BiocManager::install("RforMassSpectrometry/PSMatch")
```

To install `sager` and its dependencies, you can run the following
command:

```{r installsager, eval = FALSE}
BiocManager::install("UCLouvain-CBIO/sager")
```




# Analysing `sage` results

## Example data

```{r}
library(sager)
```

The `sager` package provides example data

```{r}
sagerAvailableData()
```

```{r}
sagerAddData()
```

For details and provenance of the data, please see the `?sagerData`
manual page.

## Identification data

```{r sagePSM, message = FALSE}
library(PSMatch)
psm <- sagePSM(sagerIdData())
psm
```

```{r ggplot, message = FALSE}
library(ggplot2)
data.frame(psm) |>
    ggplot(aes(x = hyperscore,
               colour = label)) +
    geom_density()
```

```{r ggplog2}
data.frame(psm) |>
    ggplot(aes(x = spectrum_fdr,
               colour = label)) +
    geom_density()
```

```{r filterPSM}
psm <- filterPSMs(psm) |>
    filterPsmFdr()
```

## Quantitative data

```{r, message = FALSE}
library(QFeatures)
library(scp)
```

Multiple fractions for the same set of samples:

```{r}
sageQFeatures(sagerQuantData(), sagerIdData(), splitBy = NULL)
```

Different sets of samples per acquisition:

```{r}
qf <- sageQFeatures(sagerQuantData(), sagerIdData())
qf
```

Annotate data:

```{r}
renameqf <- function(x) sub("\\.mzML", "", sub("^.+hrMS2_", "", x))
qf$filename <- rep(names(qf), each = 11)
qf$tmt_tag <- rep(paste0("tmt_", 1:11), length(qf))
qf$acquisition <- renameqf(qf$filename)
names(qf) <- paste0("psm", renameqf(names(qf)))
qf <- renamePrimary(qf, paste(qf$acquisition, qf$tmt_tag, sep = "."))
colnames(qf) <- CharacterList(lapply(colnames(qf), renameqf))
```

```{r}
qf
colData(qf)
```

Data processing:

```{r}
qf |>
    filterFeatures(~ label > 0) |>
    filterFeatures(~ rank == 1) |>
    filterFeatures(~ spectrum_fdr < 0.05) |>
    zeroIsNA(1:3) |>
    logTransform(i = 1:3, name = paste0("log_", names(qf))) |>
    aggregateFeaturesOverAssays(i = 4:6,
                                fcol = "peptide",
                                name = sub("psm", "peptide", names(qf)),
                                fun = colMedians,
                                na.rm = TRUE) |>
    joinAssays(i = 7:9,
               name = "peptides") |>
    normalize(i = 10,
              name = "norm_peptides",
              method = "center.mean") |>
    aggregateFeatures(i = "norm_peptides",
                      name = "proteins",
                      fcol = "proteins",
                      fun = MsCoreUtils::robustSummary,
                      na.rm = TRUE) -> qf
```

```{r}
qf
```

## Spectra and Quantitative data

```{r spectra, message = FALSE}
library(Spectra)
sp <- Spectra(sagerMzMLData())
sp
```

# Misc

## Asking for help

Please [open an issue](https://github.com/UCLouvain-CBIO/sager/issues)
on the package's Github repository.

<!-- ## Citing `sager` -->

<!-- ```{r citation} -->
<!-- citation("sager") -->
<!-- ``` -->

## Session information

```{r si, echo = FALSE}
sessionInfo()
```
