---
title: "Using the `sager` package to import and analyse `sage` results in R"
author:
  - name: Laurent Gatto
    affiliation:
    - Computational Biology and Bioinformatics, UCLouvain
    email: laurent.gatto@uclouvain.be
output:
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('sager')`"
vignette: >
  %\VignetteIndexEntry{The sager package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
    )
suppressPackageStartupMessages(library("BiocStyle"))
```


# Introduction

## The `sager` package

## Installation instructions

```{r "install", eval = FALSE}
if (!requireNamespace("BiocManager"))
      install.packages("BiocManager")

BiocManager::install("UCLouvain-CBIO/sager")
```

# Analysing `sage` results

## Example data

```{r}
library(sager)
```

The `sager` package provides example data

```{r}
sagerAvailableData()
```

```{r}
sagerAddData()
```

For details and provenance of the data, please see the `?sagerData`
manual page.

## Identification data

```{r}
library(PSMatch)
psm <- sagePSM(sagerIdData())
psm
```

```{r ggplot, message = FALSE}
library(ggplot2)
data.frame(psm) |>
    ggplot(aes(x = hyperscore,
               colour = label)) +
    geom_density()
```

```{r ggplog2}
data.frame(psm) |>
    ggplot(aes(x = spectrum_fdr,
               colour = label)) +
    geom_density()
```

```{r filterPSM}
psm <- filterPSMs(psm) |>
    filterPsmFdr()
```

## Quantitative data

```{r, message = FALSE}
library(QFeatures)
library(scp)
```

Multiple fractions for the same set of samples:

```{r}
sageQFeatures(sagerQuantData(), sagerIdData(), splitBy = NULL)
```

Different sets of samples per acquisition:

```{r}
qf <- sageQFeatures(sagerQuantData(), sagerIdData())
qf
```

Annotate data:

```{r}
renameqf <- function(x) sub("\\.mzML", "", sub("^.+hrMS2_", "", x))
qf$filename <- rep(names(qf), each = 11)
qf$tmt_tag <- rep(paste0("tmt_", 1:11), length(qf))
qf$acquisition <- renameqf(qf$filename)
names(qf) <- paste0("psm", renameqf(names(qf)))
qf <- renamePrimary(qf, paste(qf$acquisition, qf$tmt_tag, sep = "."))
colnames(qf) <- CharacterList(lapply(colnames(qf), renameqf))
```

```{r}
qf
colData(qf)
```

Data processing:

```{r}
qf |>
    filterFeatures(~ label > 0) |>
    filterFeatures(~ rank == 1) |>
    filterFeatures(~ spectrum_fdr < 0.05) |>
    zeroIsNA(1:3) |>
    logTransform(i = 1:3, name = paste0("log_", names(qf))) |>
    aggregateFeaturesOverAssays(i = 4:6,
                                fcol = "peptide",
                                name = sub("psm", "peptide", names(qf)),
                                fun = colMedians,
                                na.rm = TRUE) |>
    joinAssays(i = 7:9,
               name = "peptides") |>
    normalize(i = 10,
              name = "norm_peptides",
              method = "center.mean") |>
    aggregateFeatures(i = "norm_peptides",
                      name = "proteins",
                      fcol = "proteins",
                      fun = MsCoreUtils::robustSummary,
                      na.rm = TRUE) -> qf
```

```{r}
qf
```

## Spectra and Quantitative data

```{r}
library(Spectra)
sp <- Spectra(sagerMzMLData())
sp
```

# Misc

## Asking for help

Please [open an issue](https://github.com/UCLouvain-CBIO/sager/issues)
on the package's Github repository.

## Citing `sager`

```{r citation}
citation("sager")
```

## Session information

```{r si, echo = FALSE}
sessionInfo()
```
